---
title: "Big Data Ball DFS History EDA"
output:
  html_notebook: default
  html_document: default
---

## Introduction
[BigDataBall](https://www.bigdataball.com/nba-historical-dfs-stats-singleseason/) recently started collecting daily fantasy sports (DFS) data for their NBA datasets. I'm a subscriber, so I signed up for their 2016-2017 historical dataset. What follows is some preliminary exploratory data analysis.

## About the data
First, we load the spreadsheet into a tibble:

```{r warning=FALSE}
library(tidyverse)
library(readxl)
nba_dfs_2016_2017 <- read_excel(
  "~/SpiderOak Hive/NBA2016-2017/NBA-2016-2017-DFS-Dataset.xlsx",
  col_names = FALSE,
  skip = 2
)
colnames(nba_dfs_2016_2017) <- c(
  "dataset",
  "date",
  "player",
  "team",
  "opponent",
  "venue_r_h",
  "minutes",
  "usage_rate",
  "dk_position",
  "fd_position",
  "y_position",
  "dk_salary",
  "fd_salary" ,
  "y_salary",
  "dkfp",
  "fdfp",
  "yfp")
nba_dfs_2016_2017$y_salary <- as.numeric(nba_dfs_2016_2017$y_salary)
nba_dfs_2016_2017 <- nba_dfs_2016_2017 %>% filter(
  !is.na(dk_salary),
  !is.na(fd_salary),
  !is.na(y_salary),
  !is.na(dkfp),
  !is.na(fdfp),
  !is.na(yfp),
  !is.na(minutes),
  !is.na(usage_rate)
)
nba_dfs_2016_2017
```

We have 17 columns. Most of the columns are self-explaining, but one may be new to people other than basketball analytics fans - "usage".

## Outlier removal
It turns out that there are some outliers in the usage calculation. Since usage has minutes played, a player who clocked a low number of minutes but ran up some points will have a high usage value. So I used a boxplot to find those outliers.
```{r}
boxplot_stats <- boxplot(nba_dfs_2016_2017$usage_rate)$stats
nba_dfs_2016_2017 <- nba_dfs_2016_2017 %>% 
  filter(usage_rate <= boxplot_stats[5, 1]) 
```


## Clustering
The first step in my exploration process on a multivariate dataset like this one is clustering. There are a few ways to do this; it's a popular unsupervised learning technique to "let the data speak for themselves". My default clustering EDA library package is "Rmixmod".

```{r}
cluster_input <- nba_dfs_2016_2017 %>%
  select(minutes, usage_rate, dk_salary, dkfp)
library(Rmixmod)
clusters <- mixmodCluster(cluster_input, nbCluster = 1:3)
print(summary(clusters))
plot(clusters, main = "DraftKings")
```

