---
title: "Big Data Ball DFS History EDA"
output:
  html_notebook: default
  html_document: default
---

## Introduction
[BigDataBall](https://www.bigdataball.com/nba-historical-dfs-stats-singleseason/) recently started collecting daily fantasy sports (DFS) data for their NBA datasets. I'm a subscriber, so I signed up for their 2016-2017 historical dataset. What follows is some preliminary exploratory data analysis.

## About the data
First, we load the spreadsheet into a tibble:

```{r warning=FALSE}
library(tidyverse)
library(readxl)
nba_dfs_2016_2017 <- read_excel(
  "~/SpiderOak Hive/NBA2016-2017/NBA-2016-2017-DFS-Dataset.xlsx",
  col_names = FALSE,
  skip = 2
)
colnames(nba_dfs_2016_2017) <- c(
  "dataset",
  "date",
  "player",
  "team",
  "opponent",
  "venue_r_h",
  "minutes",
  "usage_rate",
  "dk_position",
  "fd_position",
  "y_position",
  "dk_salary",
  "fd_salary" ,
  "y_salary",
  "dkfp",
  "fdfp",
  "yfp")
nba_dfs_2016_2017$y_salary <- as.numeric(nba_dfs_2016_2017$y_salary)
nba_dfs_2016_2017 <- nba_dfs_2016_2017 %>% filter(
  !is.na(dk_salary),
  !is.na(fd_salary),
  !is.na(y_salary),
  !is.na(dkfp),
  !is.na(fdfp),
  !is.na(yfp),
  !is.na(minutes),
  !is.na(usage_rate)
)
nba_dfs_2016_2017
```

## Usage
We have 17 columns. Most of the columns are self-explaining, but one may be new to people other than basketball analytics fans - "usage".

## Scatterplot matrix
The first order of business in this exploration is to get a display of the characteristics of the numbers in this dataset. For this, we'll use a scatterplot matrix from the "GGally" package. To simplify matters, we'll only look at DraftKings DFS data.
```{r fig.width=11, fig.height=8.5}
library(GGally)
ggpairs(
  data = nba_dfs_2016_2017 %>% select(
    dk_salary,
    minutes, 
    usage_rate,
    dkfp
  )
)
```

Notice that the "usage_rate" ranges from zero up to past 1500! The large values are clearly outliers. Where are they coming from?

As you can see above, the formula for "usage_rate" has minutes played in the denominator. So if a player makes any signigicant contribution in a small time played, they'll have an unreasonable large usage rate.

Notice also that there's an abnormally large spike in minutes around five. We want to filter these out as well as the large usage_rate values. So we'll remove all rows with less than ten minutes played.

```{r fig.width=11, fig.height=8.5}
no_outliers <- nba_dfs_2016_2017 %>% filter(
  minutes >= 10
)
ggpairs(
  data = no_outliers %>% select(
    dk_salary,
    minutes, 
    usage_rate,
    dkfp
  )
)
```

## Clustering
The first step in my exploration process on a multivariate dataset like this one is clustering. There are a few ways to do this; it's a popular unsupervised learning technique to "let the data speak for themselves". My default clustering EDA library package is "Rmixmod".

```{r fig.width=11, fig.height=8.5}
cluster_input <- no_outliers %>%
  select(minutes, usage_rate, dk_salary, dkfp)
library(Rmixmod)
clusters <- mixmodCluster(cluster_input, nbCluster = 3)
print(summary(clusters))
plot(clusters)
```

